---
title: A closer look at replicate() and purrr::map() for simulations
author: Ariel Muldoon
date: '2018-06-05'
slug: a-closer-look-at-replicate-and-purrr
categories:
  - r
tags:
  - purrr
  - simulation
draft: FALSE
description: "In this post I delve into the details of the R functions I've been using in my simulation examples, focusing on the replicate() function and the map family of functions from the purrr package.  I spend a little time showing the parallels between the replicate() function and a for() loop."
---

```{r setup, echo = FALSE}
knitr::opts_chunk$set(comment = "#")
```

I've done a couple of posts so far on simulations, [here](https://aosmith.rbind.io/2018/01/09/simulate-simulate-part1/) and [here](https://aosmith.rbind.io/2018/04/23/simulate-simulate-part-2/), where I demonstrate how to build a function for simulating data from a defined linear model and then explore long-run behavior of models fit to the simulated datasets.  The focus of those posts was on the general simulation process, and I didn't go into much detail on the specific R code.  In this post I'll focus in on the code I use for repeatedly simulating data and extracting output, specifically talking about the function `replicate()` and the *map* family of functions from package **purrr**.

# The replicate() function

The `replicate()` function is a member of the *apply* family of functions in base R.  
Specifically, from the documentation:

> `replicate` is a wrapper for the common use of `sapply` for repeated evaluation of an expression (which will usually involve random number generation).

Notice the documentation mentions *repeated evaluations* and that the use of `replicate()` involves *random number generation*.  Those are primary parts of the simulations I do.  While I don't actually know the *apply* family of functions very well, I use `replicate()` a lot (although also see `purrr::rerun()`).  Using `replicate()` is an alternative to building a `for()` loop to repeatedly simulate new values.

The `replicate()` function takes three arguments:

* `n`, which is the number of replications to perform.  This is where I set the number of simulations I want to run.  
* `expr`, the expression that should be run repeatedly.  I've only ever used a function here.  
* `simplify`, which controls the type of output the results of `expr` are saved into.  Use `simplify = FALSE` to get vectors saved into a list instead of in an array.  

## Simple example of replicate()

Let's say I wanted to simulate some values from a normal distribution, which I can do using the `rnorm()` function.  Below I'll simulate five values from a normal distribution with a mean of 0 and a standard deviation of 1 (which are the defaults for `mean` and `sd` arguments, respectively).

Since I'm going generate random numbers I'll set the seed so anyone following along at home will see the same values.

```{r}
set.seed(16)
rnorm(5, mean = 0, sd = 1)
```

Using `rnorm()` directly gives me a single set of simulated values.  How do I simulate 5 values from this same distribution multiple times?  This is where `replicate()` comes in.  It allows me to run the function I put in `expr` exactly `n` times.  

Here I'll ask for three runs of 5 values each.  Notice I use `simplify = FALSE` to get a list as output.

The output below is a list of three vectors.  Each vector is from a unique run of the function, so contains  five random numbers drawn from the normal distribution with a mean of 0 and standard deviation of 1.

```{r}
set.seed(16)
replicate(n = 3, rnorm(5, 0, 1), simplify = FALSE )
```

Note if I don't use `simplify = FALSE` I will get a matrix of values instead of a list.  Each column in the matrix is the output from one run of the function.  In this case there will be three columns in the output, one for each run, and 5 rows.  This can be a useful output type for simulations.  I focus on list output throughout the rest of this post only because that's what I have been using recently for simulations.

```{r}
set.seed(16)
replicate(n = 3, rnorm(5, 0, 1) )
```

## An equivalent for() loop example

A `for()` loop can be used in place of `replicate()` for simulations.  With time and practice I've found `replicate()` to be much more convenient in terms of writing the code.  However, in my experience some folks find `for()` loops intuitive when they are starting out in R.  I think it's because `for()` loops are more explicit on the looping process: the user can see that `i` is looped over and the output for each `i` iteration is saved into the output object because the code is written out explicitly.

In my example I'll save the output of each iteration of the loop into a list called `list1`.  I initialize this as an empty list prior to starting the loop.  To match what I did with `replicate()` I do three iterations of the loop (`i in 1:3`), drawing 5 values via `rnorm()` each time.

The result is identical to my `replicate()` code above.  It took a little more code to do it but the process is very clear since it is explicitly written out.

```{r}
set.seed(16)
list1 = list() # Make an empty list to save output in
for (i in 1:3) { # Indicate number of iterations with "i"
    list1[[i]] = rnorm(5, 0, 1) # Save output in list for each iteration
}
list1
```

## Using replicate() on a user-made function

When I do simulations to explore the behavior of linear models under different scenarios I will create a function to simulate the data and fit the model.  For example, here's a function I used in an [earlier blog post](https://aosmith.rbind.io/2018/01/09/simulate-simulate-part1/) to simulate data from and then fit a two group linear model.

```{r}
twogroup_fun = function(nrep = 10, b0 = 5, b1 = -2, sigma = 2) {
     ngroup = 2
     group = rep( c("group1", "group2"), each = nrep)
     eps = rnorm(ngroup*nrep, 0, sigma)
     growth = b0 + b1*(group == "group2") + eps
     growthfit = lm(growth ~ group)
     growthfit
}
```

The output is a model fit to data generated from the fixed and random (residual) effects.

```{r}
twogroup_fun()
```

To explore the long-run behavior in my simulated scenario I will repeat the data generation and model fitting many times using `replicate()`.  The result is a list of fitted models.  I'll run the function 5 times and save the result as `sim_lm` to use throughout the next section on `map()`.

```{r}
sim_lm = replicate(5, twogroup_fun(), simplify = FALSE )
length(sim_lm)
```

# Using purrr::map() for looping through lists

So I have a list of fitted models from `replicate()`; now what?   

The `replicate()` function was about repeatedly running a function.  Once I have the repeated runs I can explore the long-run behavior of some statistic by extracting value(s) from the resulting models.  This involves looping through the list of models.  

Looping through the list can be done using a `for()` loop, but I prefer to use functions that do the looping without all the typing.  In particular, these days I use the *map* family of functions from the **purrr** package to loop through lists.  Before **purrr** I primarily used `lapply()` (the only other *apply* family function that I know `r emo::ji("laughing")`). 

The `map()` function takes a list as input and puts the output into a list of the same length. The first argument to `map()` is the list to loop through and the second argument is the function to apply to each element of the list.  

For example, I can pull out the coefficients of each model in my 5-run simulation by looping through `sim_lm` and applying the `coef()` function to each list element.

```{r}
library(purrr) # v. 0.2.5
map(sim_lm, coef)
```

## Other variants of map() for non-list outputs

There are many variants of `map()` that are convenient for saving results into something other than a list.  For example, if I am going to extract a single numeric value from each model, such as $R^2$, I might want the output to be a numeric vector instead of a list.  I can use `map_dbl()` for this.

The unadjusted $R^2$ from a model fit with `lm()` can be pulled from the model `summary()` output.  The code looks like: `summary(model)$r.squared`,   
where "model" is a fitted model.  

So getting $R^2$ involves extracting a value after applying a function, which isn't quite as straightforward as applying a single function to every model in the list like I did with `coef()`.  This gives me a chance to demonstrate the formula coding styling available in *map* functions.  In formula coding a tilde (`~`) goes in front of the function and `.x` refers to the list element.

```{r}
map_dbl(sim_lm, ~summary(.x)$r.squared)
```

If you don't like the formula style you can use an anonymous function inside *map* functions, where the function argument is used to refer to the list element.  

```{r}
map_dbl(sim_lm, function(x) summary(x)$r.squared)
```

For data.frame output we can use `map_dfr()` for row binding or *stacking* results together into a single data.frame.

Estimated coefficients, their standard errors, and their statistical tests from models fit with `lm()` can be extracted into a tidy data.frame using `broom::tidy()`.  Looping through the results and doing this for each model via `map_dfr()` will put the output in one data.frame instead of storing the individual data.frames for each model as one element of a list.

```{r}
map_dfr(sim_lm, broom::tidy)
```

The `map_dfr()` function has an additional argument, `.id`, which can be used to store the list names (if the original list had names) or add the list index to the output (if it didn't have names).  I'm using a list that has no names, so each unique model output will be assigned its index number if I use the `.id` argument.  The name of the new column is given as a string to `.id`.

```{r}
map_dfr(sim_lm, broom::tidy, .id = "model")
```

Further arguments to the function used within `map()` can be passed as additional arguments.  For example, I can add confidence intervals for estimated coefficients when using the `tidy.lm()` function via `conf.int = TRUE`.  If I want to get confidence intervals for all models I add this as an additional argument in `map_dfr()`.

```{r}
map_dfr(sim_lm, broom::tidy, conf.int = TRUE)
```

The *map* family of functions can easily be used with pipes as one step in a chain of functions.  I can, for example, take the estimates I get using `broom::tidy`, pull out the estimated intercepts, and then plot a histogram of those estimates.  I'll need packages **dplyr** and **ggplot2** for this.

```{r}
suppressPackageStartupMessages( library(dplyr) ) # v. 0.7.5
library(ggplot2) # v 2.2.1
```

You can see all the steps in the pipe chain below.  I loop through `sim_lm` using `map_dfr()` to extract the coefficients from each element of the list and output a data.frame of results.  I use `dplyr::filter()` to keep only the rows with estimated intercepts and then plot a histogram of these estimates for the whole simulation with `ggplot2::qplot()`.

```{r}
sim_lm %>%
     map_dfr(broom::tidy) %>%
     filter(term == "(Intercept)") %>%
     qplot(x = estimate, data = ., geom = "histogram")
```

There are more variants of `map()` that you might find useful, both for simulations and in other iterative work.  See the documentation for `map()` (`?map`) to see all of them along with additional examples.
