---
title: 'Time after time: temporal autocorrelation with independent groups'
author: Ariel Muldoon
date: '2018-06-21'
slug: group-autocorrelation
categories:
  - r
  - statistics
tags:
  - autocorrelation
  - tidyr
draft: TRUE
description: "Autocorrelation can be underestimated When independent groups are ignored.  I show an approach to respect groups by padding the dataset with NA via tidyr::complete()."
---

Simulate data with autocorrelated noise via `arima.sim()`.  

```{r}
library(purrr)
suppressPackageStartupMessages( library(dplyr) )

dat = map2_dfr(1:10, sample(6:9, 10, replace = TRUE), 
               ~tibble(group = .x,
                     x = runif(.y, 5, 10),
                     y = 1 + x + arima.sim(list(ar = .7), .y),
                     time = 1:.y)
)
```

Check group sizes

```{r}
dat %>%
    count(group) %>%
    filter(n == max(n) )
```

Fit model and check for autocorrelation in the residuals via `acf()`.  I'm using a maximum lag is based on the smallest group size.

```{r}
fit1 = lm(y ~ x, data = dat)

dat$res = residuals(fit1)

acf(dat$res, plot = FALSE, lag.max = 6)
```

Expand dataset using `tidyr::complete()`.  A variable representing the autocorrelation variable is needed.  In this case it's `time`, but if I was doing soil depth that was taken evenly through a soil core it could be something like soild depth.

I group the dataset so every group is padded with rows of missing values at the end.

```{r}
dat_expand = dat %>%
    group_by(group) %>%
    complete(time = 1:20) 
```

Here is an example of the first group, which has X observations and then 10 rows of `NA`.

```{r}
filter(dat_expand, group == 1)
```

Now calculate the residual autocorrelation again using the expanded dataset, using `na.action = na.pass`.  Now groups are not mistakenly considered in the autocorrelation since rows of observations are no longer contiguous.

```{r}
acf(dat_expand$res, lag.max = 7, plot = FALSE, na.action = na.pass)
```

For linear models, either with or without random effects, the **nlme** package as an `ACF` function that can be calculated by group. The calculation is very slightly different than the one we did, but clearly shows how I had underestimated the autocorrelation originally.

```{r}
library(nlme)
fit2 = gls(y ~ x, data = dat)
ACF(fit2, maxLag = 7, form = ~1|group)
```

