---
title: Custom contrasts in emmeans
author: Ariel Muldoon
date: '2019-04-15'
slug: custom-contrasts-emmeans
categories:
  - r
  - statistics
tags:
  - analysis
  - teaching
  - emmeans
draft: TRUE
description: "One of the nice things about emmeans is that you can build custom comparisons for any groups or combinations of groups, which I show how to do in this post."
---


```{r setup, echo = FALSE}
knitr::opts_chunk$set(comment = "#")
```

```{r, echo = FALSE}
# Make dataset here that will use later
# Two factors plus one log-normal response
set.seed(16)
dat = data.frame(f1.f2 = rep(c("A.1", "A.2", "B.1", "B.2", "control"), each = 5) )
y = with(dat, 5 + 
              -1*(f1.f2 == "A.2") +
              .5*(f1.f2 == "B.1") +
              1*(f1.f2 == "B.2") +
              2*(f1.f2 == "control") +
              rnorm(n = 25, mean = 0, sd = 1) )
dat$resp = round(y, 1)
```

Following up on a [previous post](https://aosmith.rbind.io/2019/03/25/getting-started-with-emmeans/) where I demonstrated the basic usage of package **emmeans** for doing post hoc comparisons, here I'll demonstrate how to make custom comparisons (aka *contrasts*).  These are comparisons that aren't encompassed by the built-in functions in the package. 

Remember that you can explore the available built-in **emmeans** functions for doing comparisons via `?"contrast-methods"`.

# Reasons for custom comparisons

There are a variety of reasons you might need custom comparisons instead of some of the standard, built-in ones.  One common scenario that I see a lot is when we have a single control group for multiple factors, so the factors aren't perfectly crossed.  This comes up, e.g., when doing experiments that involve applying different substances (like fertilizers) at varying rates.  One factor is the different substances applied and the other is different application rates.  However, the control is applying nothing or water or something like that.  There aren't different "rates" of the control to apply, so there is a single control group for both factors.

Rather than trying to fit a model with multiple factors, focusing on main effects and the interaction, such data can be analyzed with a *simple effects* model.  This is where the two (or more) factors of interest have been combined into a single factor for analysis.  This analysis focuses on the effect of the two factors combined.  We can use post hoc comparisons to estimate the overall effects of individual factors.  

# R packages

I will load **magrittr** for the pipe in addition to **emmeans**.

```{r}
library(emmeans) # v. 1.3.3
library(magrittr) # v. 1.5
```

# The dataset and model

I've made a small dataset to use as an example.  The response variable is `resp` and the two factors of interest have been combined into a single factor `f1.f2` that has 5 levels: `A.1`, `A.2`, `B.1`, `B.2`, and `control`.  

One factor, which I'm thinking of as the "substance" factor, is represented by `A` and `B` (and the control).  The second, the "rate" factor, is represented by `1` and `2`.

```{r}
dat = structure(list(f1.f2 = structure(c(1L, 1L, 1L, 1L, 1L, 2L, 2L, 
2L, 2L, 2L, 3L, 3L, 3L, 3L, 3L, 4L, 4L, 4L, 4L, 4L, 5L, 5L, 5L, 
5L, 5L), .Label = c("A.1", "A.2", "B.1", "B.2", "control"), class = "factor"), 
    resp = c(5.5, 4.9, 6.1, 3.6, 6.1, 3.5, 3, 4.1, 5, 4.6, 7.3, 
    5.6, 4.8, 7.2, 6.2, 4.3, 6.6, 6.5, 5.5, 7.1, 5.4, 6.7, 6.8, 
    8.5, 6.1)), row.names = c(NA, -25L), class = "data.frame")
str(dat)
```

I will use a simple, linear model for analysis.

```{r}
fit1 = lm(resp ~ f1.f2, data = dat)
```

# Treatment vs control comparisons

This simple effects model makes it easy to get comparisons for each factor combination vs the control group with `emmeans()`.  I'll use `trt.vs.ctrlk` to do this since the control is the last level of the factor.

```{r}
emmeans(fit1, specs = trt.vs.ctrlk ~ f1.f2)
```

# [Getting started building custom contrasts](#getting-started-building-custom-contrasts)

But what if we want to compare the `A` group to the `B` group overall, regardless of the application rate?  There is no built-in way to do this; we'll need to set up a custom contrast.

Custom contrasts are based on the estimated marginal means output from `emmeans()`.  The first step is to calculate these so we have them to work with.  I name this output `emm1`.

```{r}
emm1 = emmeans(fit1, specs = ~ f1.f2)
emm1
```

Before I get to that more complicated example of `A` vs `B` overall, I'm going to walk through a simpler example.  I'll start by showing how to compare mean `resp` of the `A.2` group to the `B.2` group via custom contrasts.  This involves pulling out those specific group means from the `emmeans` output shown above.  

We *pull out* a group mean by making a vector to represent the specific mean of interest.  In this vector we assign a `1` to the mean of the group of interest and a `0` to the other groups.  

For example, to pull out the mean of `A.2` we will make a vector with 5 values in it, one for each row of the `emmeans` output.  The second value will be a `1`, since the mean of `A.2` is on the second row.  All the other values in the vector will be `0`.

Below is the vector that represents that `A.2` mean.

```{r}
A2 = c(0, 1, 0, 0, 0)
```

Similarly, to pull out the mean of `B.2` we'll have a vector of 5 values with a `1` as the fourth value.  In the output above the `B.2` group is on the fourth row.

```{r}
B2 = c(0, 0, 0, 1, 0)
```

When building custom contrasts via vectors like this, the vectors will always be the same length as the number of rows in the `emmeans` output.  I always calculate and print the `emmeans` prior to building the vectors so I am certain of the number of rows and the order of the groups in the output.

# Using the contrast() function for custom comparisons

Once we have the vectors that represent the means we are interested in comparing, we actually do the comparisons via the `contrast()` function.  Since we are interested in a *difference* in mean response, we take the difference between the vectors that represent the means.  Taking the difference can be done inside or outside `contrast()`.  In this example I'm doing it inside.

The `contrast()` function takes an `emmGrid` object (i.e., output from `emmeans()`) as the first argument.  We give the comparison we want to do via a list passed to the `method` argument.  

Here I want to calculate the difference in mean `resp` of `A.2` and `B.2`.  I subtract the `B2` vector from the `A2` vector.

The output is the difference in mean `resp` between these groups.

```{r}
contrast(emm1, method = list(A2 - B2) )
```

# Using named lists for better output

The name in the output above isn't very helpful for figuring out which groups this comparison represents.  Using a named list for `method` is much more useful.

```{r}
contrast(emm1, method = list("A2 - B2" = A2 - B2) )
```

# [Using "at" for simple comparisons](#using-at-for-simple-comparisons)

Note that I didn't need to do a custom contrast to do this particular comparison.  I could have gotten the comparison I wanted by using the `at` argument in `emmeans()` and choosing just the two groups I was interested in.

```{r}
emmeans(fit1, specs = pairwise ~ f1.f2, 
         at = list(f1.f2 = c("A.2", "B.2") ) )
```

# [Multiple custom contrasts at once](#multiple-custom-contrasts-at-once)

Multiple custom contrasts can be done simultaneously by adding more comparisons to the `method` list.  I'll demonstrate this by doing the simple example comparison twice, changing only which group is subtracted from the other.  I name both elements of the list for ease of interpreting the output.  

```{r}
contrast(emm1, method = list("A2 - B2" = A2 - B2,
                             "B2 - A2" = B2 - A2) )
```

Doing multiple comparisons at once means a multiple comparisons adjustment can be done as needed.  In addition, we can use the `confint()` function do get confidence intervals for the comparisons.

I'll play around here by adding a multivariate-$t$ adjustment via `adjust = "mvt"` and then get confidence intervals for the comparisons.  Remember we can get both confidence intervals and tests for comparisons via `summary()` with `infer = TRUE`.

```{r}
contrast(emm1, method = list("A2 minus B2" = A2 - B2,
                             "B2 minus A2" = B2 - A2),
         adjust = "mvt") %>%
     confint()
```

# More complicated custom contrasts

Let's go back to the original comparison I proposed, which was to estimate the overall effect of `A` vs `B`.  This is a *main effect* comparison, so I need to average over the effect of the "rate" factor in order to estimate the overall effect of the "substance" factor.

To do this comparison I need the means for all four non-control factor levels.  I'll print `emm1` again here so I remember the order of the output.

```{r}
emm1
```

I pull out each of the first four group means, just like I demonstrated above.  While typing these vectors out isn't too hard, since they only contain 5 values, if I have many groups and so really long vectors you will sometimes see me use `rep()` to put all the 0 values in.

```{r}
A1 = c(1, 0, 0, 0, 0)
A2 = c(0, 1, 0, 0, 0)
B1 = c(0, 0, 1, 0, 0)
B2 = c(0, 0, 0, 1, 0)
```

The vectors I made represent means for combinations of substance and rate.  I want the overall substance group means, though.  This is done by *averaging over* the two rates.  I literally taking the average of, e.g., `A1` and `A2` vectors to get a vector that represents the overall `A` mean.

```{r}
Aoverall = (A1 + A2)/2
Boverall = (B1 + B2)/2
```

Now that the overall means are calculated we can do comparison of mean `resp` of the `A` group vs `B` group overall in `contrast()`.

```{r}
contrast(emm1, method = list("A - B" = Aoverall - Boverall) ) 
```

Custom contrasts are all built in this same basic way.  You can also build your own contrast function if there is some contrast you do all the time that is not part of **emmeans**.  See the [custom contrasts section](https://cran.r-project.org/web/packages/emmeans/vignettes/comparisons.html#linfcns) of the **emmeans** vignette for more info.
