---
title: 'The small multiples plot: how to combine ggplot2 plots with one shared axis'
author: Ariel Muldoon
date: '2019-05-13'
slug: small-multiples-plot
categories:
  - r
tags:
  - ggplot2
  - tidyr
draft: FALSE
description: "There are a variety of ways to combine ggplot2 plots with a single shared axis, but things can get tricky if you want a lot of control over all plot elements.  I show three approaches to make such a plot: using facets, with package cowplot, and with package egg."
---

```{r setup, echo = FALSE, message = FALSE}
knitr::opts_chunk$set(comment = "#")
devtools::source_gist("c83e078bf8c81b035e32c3fc0cf04ee8", 
                      filename = 'render_toc.R')
```

There are a variety of ways to combine **ggplot2** plots with a single shared axis.  However, things can get tricky if you want a lot of control over all plot elements.  

I demonstrate three different approaches for this:  
1. Using facets, which is built in to **ggplot2** but doesn't allow much control over the non-shared axes.  
2. Using package **cowplot**, which has a lot of nice features but the plot spacing doesn't play well with a single shared axis.  
3. Using package **egg**, which allows nice spacing for plots with a shared axis.  

## Table of Contents

```{r toc, echo = FALSE} 
render_toc("2019-05-13-small-multiples-plot.Rmd")
```

# Load R packages

I'll be plotting with **ggplot2**, reshaping with **tidyr**, and combining plots with package **egg**.  

I'll also be using package **cowplot** version 0.9.4 to combine individual plots into one, but will use the package functions via `cowplot::` instead of loading the package.  (I believe the next version of **cowplot** will not be so opinionated about the theme.)

```{r, message = FALSE}
library(ggplot2) # v. 3.1.1
library(tidyr) # v. 0.8.3
library(egg) # v. 0.4.2
```

# The set-up

Here's the scenario: we have one response variable (`resp`) that we want to plot against three other variables and combine them into a single "small multiples" plot.  

I'll call the three variables `elev`, `grad`, and `slp`.  You'll note that I created these variables to have very different scales.

```{r}
set.seed(16)
dat = data.frame(elev = round( runif(20, 100, 500), 1),
                 resp = round( runif(20, 0, 10), 1),
                 grad = round( runif(20, 0, 1), 2),
                 slp = round( runif(20, 0, 35),1) )
head(dat)
```

# Using facets for small multiples

One good option when we want to make a similar plot for different groups (in this case, different variables) is to use *faceting* to make different panels within the same plot.  

Since the three variables are currently in separate columns we'll need to *reshape* the dataset prior to plotting.  I'll use `gather()` from **tidyr** for this.

```{r}
datlong = gather(dat, key = "variable", value = "value", -resp)
head(datlong)
```

Now I can use `facet_wrap()` to make a separate scatterplot of `resp` vs each variable.  The argument `scales = "free_x"` allows the x axis scales to differ for each variable but leaves a single y axis.

```{r}
ggplot(datlong, aes(x = value, y = resp) ) +
     geom_point() +
     theme_bw() +
     facet_wrap(~variable, scales = "free_x")
```

I can use the facet strips to give the appearance of axis labels, as shown in [this Stack Overflow answer](https://stackoverflow.com/a/37574221/2461552).

```{r}
ggplot(datlong, aes(x = value, y = resp) ) +
     geom_point() +
     theme_bw() +
     facet_wrap(~variable, scales = "free_x", strip.position = "bottom") +
     theme(strip.background = element_blank(),
           strip.placement = "outside") +
     labs(x = NULL)
```

That's a pretty nice plot to start with.  However, controlling the axis breaks in the individual panels [can be complicated](https://stackoverflow.com/questions/51735481/ggplot2-change-axis-limits-for-each-individual-facet-panel), which is something we'd commonly want to do.  

In that case, it may make more sense to create separate plots and then combine them into a small multiples plot with an add-on package.

# Using **cowplot** to combine plots

Package **cowplot** is a really nice package for combining plots, and has lots of bells and whistles along with some pretty thorough [vignettes](https://cran.r-project.org/web/packages/cowplot/index.html).

The first step is to make each of the three plots separately.  If doing lots of these we'd want to use some sort of loop to make a list of plots [as I've demonstrated previously](https://aosmith.rbind.io/2018/08/20/automating-exploratory-plots/).  Today I'm going to make the three plots manually.

```{r}
elevplot = ggplot(dat, aes(x = elev, y = resp) ) +
     geom_point() +
     theme_bw()

gradplot = ggplot(dat, aes(x = grad, y = resp) ) +
     geom_point() +
     theme_bw() +
     scale_x_continuous(breaks = seq(0, 1, by = 0.2) )

slpplot = ggplot(dat, aes(x = slp, y = resp) ) +
     geom_point() +
     theme_bw() +
     scale_x_continuous(breaks = seq(0, 35, by = 5) )
```

The function `plot_grid()` in **cowplot** is for combining plots. To make a single row of plots I use `nrow = 1`.  

The `labels` argument puts separate labels on each panel for captioning.

```{r}
cowplot::plot_grid(elevplot, 
                   gradplot, 
                   slpplot,
                   nrow = 1,
                   labels = "auto")
```

But we want a single shared y axis, not a separate y axis on each plot.  I'll remake the combined plot, this time removing the y axis elements from all but the first plot.

```{r}
cowplot::plot_grid(elevplot, 
                   gradplot + theme(axis.text.y = element_blank(),
                                    axis.ticks.y = element_blank(),
                                    axis.title.y = element_blank() ), 
                   slpplot + theme(axis.text.y = element_blank(),
                                    axis.ticks.y = element_blank(),
                                    axis.title.y = element_blank() ),
                   nrow = 1,
                   labels = "auto")
```

This makes panels different sizes, though, which isn't ideal.  To have all the plots the same width I need to align them vertically with `align = "v"`.

```{r}
cowplot::plot_grid(elevplot, 
                   gradplot + 
                        theme(axis.text.y = element_blank(),
                              axis.ticks.y = element_blank(),
                              axis.title.y = element_blank() ), 
                   slpplot + 
                        theme(axis.text.y = element_blank(),
                              axis.ticks.y = element_blank(),
                              axis.title.y = element_blank() ),
                   nrow = 1,
                   labels = "auto",
                   align = "v")
```

But, unfortunately, this puts the axis space back between the plots to make them all the same width.  It turns out that **cowplot** isn't really made for plots with a single shared axis.  The **cowplot** package author points us to package **egg** for this [in this Stack Overflow answer](https://stackoverflow.com/a/47615304/2461552).

# Using **egg** to combine plots

Package **egg** is another nice alternative for combining plots into a small multiples plot.  The function in this package for combining plots is called `ggarrange()`.

Here are the three plots again.  

```{r}
elevplot = ggplot(dat, aes(x = elev, y = resp) ) +
     geom_point() +
     theme_bw()

gradplot = ggplot(dat, aes(x = grad, y = resp) ) +
     geom_point() +
     theme_bw() +
     scale_x_continuous(breaks = seq(0, 1, by = 0.2) )

slpplot = ggplot(dat, aes(x = slp, y = resp) ) +
     geom_point() +
     theme_bw() +
     scale_x_continuous(breaks = seq(0, 35, by = 5) )
```

The `ggarrange()` function has an `nrow` argument so I can keep the plots in a single row.  

The panel spacing is automagically the same here after I remove the y axis elements, and things look pretty nice right out of the box.

```{r}
ggarrange(elevplot, 
          gradplot + 
               theme(axis.text.y = element_blank(),
                     axis.ticks.y = element_blank(),
                     axis.title.y = element_blank() ), 
          slpplot + 
               theme(axis.text.y = element_blank(),
                     axis.ticks.y = element_blank(),
                     axis.title.y = element_blank() ),
          nrow = 1)
```

We can bring panes closer by removing some of the space around the plot margins with the `plot.margin` in `theme()`.  I'll set the spacing for right margin of the first plot, both left and right margins of the second, and the left margin of the third.

```{r}
ggarrange(elevplot +
               theme(axis.ticks.y = element_blank(),
                     plot.margin = margin(r = 1) ), 
          gradplot + 
               theme(axis.text.y = element_blank(),
                     axis.ticks.y = element_blank(),
                     axis.title.y = element_blank(),
                     plot.margin = margin(r = 1, l = 1) ), 
          slpplot + 
               theme(axis.text.y = element_blank(),
                     axis.ticks.y = element_blank(),
                     axis.title.y = element_blank(),
                     plot.margin = margin(l = 1)  ),
          nrow = 1)
```

# Adding plot labels with tag_facet() in **egg**

You'll see there is a `labels` argument in `ggarrange()` documentation, but it didn't work well for me out of the box with only one plot with a y axis.  However, we can get tricky with `egg::tag_facet()` if we add a facet strip to each of the individual plots.

It'd make sense to build these plots outside of `ggarrange()` and then add the tags and combine them instead of nesting everything like I did here, since the code is now a little hard to follow.

```{r}
ggarrange(tag_facet(elevplot +
                          theme(axis.ticks.y = element_blank(),
                                plot.margin = margin(r = 1) ) +
                          facet_wrap(~"elev"),
                     tag_pool = "a"), 
          tag_facet(gradplot + 
                          theme(axis.text.y = element_blank(),
                                axis.ticks.y = element_blank(),
                                axis.title.y = element_blank(),
                                plot.margin = margin(r = 1, l = 1) ) +
                          facet_wrap(~"grad"), 
                    tag_pool = "b" ), 
          tag_facet(slpplot + 
                          theme(axis.text.y = element_blank(),
                                axis.ticks.y = element_blank(),
                                axis.title.y = element_blank(),
                                plot.margin = margin(l = 1)  ) +
                          facet_wrap(~"slp"),
                     tag_pool = "c"),
          nrow = 1)
```

We might want to add a right y axis to the right-most plot.  In this case we'd want to change the axis ticks length to 0 via `theme()` elements.  This can be done [separately per axis in the development version of **ggplot2**](https://github.com/tidyverse/ggplot2/pull/2934), and will be included in version 3.2.0.
