---
title: Controlling legend appearance in ggplot2 with override.aes
author: Ariel Muldoon
date: '2020-07-09'
slug: ggplot2-override-aes
categories:
  - r
tags:
  - ggplot2
description: Changing the legend appearance without changing the plot appearance can be done using the override.aes argument to guide_legend().  I go through examples of when this is useful, starting with a basic example making a small legend change and working up towards using override.aes to make a combined legend.
draft: TRUE
---

```{r setup, include = FALSE, message = FALSE, purl = FALSE}
knitr::opts_chunk$set(comment = "#")
devtools::source_gist("2500a85297b742c6f2fb3a14549f5851",
                      filename = 'render_toc.R')
```

In **ggplot2**, aesthetics and their `scale_*()` functions change both the plot and the plot legend appearance simultaneously.  The `override.aes` argument in `guide_legend()` allows us to change only the legend appearance without affecting the rest of the plot.  This can allow us to make the legend more readable or make combined legends when mapping factors to different aesthetics.  In this post I'll introduce the basic usage of `override.aes` and then go through some examples of when it might come in handy.

## Table of Contents

```{r toc, echo = FALSE, purl = FALSE} 
input  = knitr::current_input()
render_toc(input)
```

# R packages

The only package I'm using is **ggplot2**.

```{r, warning = FALSE, message = FALSE}
library(ggplot2) # v. 3.3.2

```

# Using override.aes

One basic reason we might want to change the legend without changing the plot is to make the legend more readable.  

For example, I'll start with a scatterplot using the `diamonds` dataset.  This is a large dataset, so after mapping `color` to the `cut` variable I set `alpha` so the points are transparaent and reduce the `size` of points.

You can see these changes in both the plot and the legend.

```{r firstplot}
ggplot(data = diamonds, aes(x = carat, y = price, color = cut) ) +
     geom_point(alpha = .25, size = 1)
     
```


## Adding a `guides()` layer

Making the points small and transparent make the legend more difficult to read.  This is a case where I'd want to change the legend appearance to make it more readable by increasing the point size and/or reducing the transparency.

One way to do this is to add a `guides()` layer.  This uses scale name-guide pairs.  I am going to change the legend for the `color` scale, so I'll use `color = guide_legend()` as the scale name-guide pair.  

`override.aes` is an argument within `guide_legend()`, so if you're looking for more background then `?guide_legend` is a good place to start.  The argument takes a list of aesthetic parameters that will *override* the default legend appearance.  

If I want to increase the `size` of the points in the `color` legend of my plot, the layer I'll add will look like:

```{r, eval = FALSE}
guides(color = guide_legend(override.aes = list(size = 3) ) )
```

Let's add that to the plot so you can see how the points in the legend get larger while the points in the plot remain unchanged.

```{r firstplot2}
ggplot(data = diamonds, aes(x = carat, y = price, color = cut) ) +
     geom_point(alpha = .25, size = 1) +
     guides(color = guide_legend(override.aes = list(size = 3) ) )
            
```

## Using the `guide` argument

If I was going to change my default colors with a `scale_color_*()` function, I could use the `guide` argument there to override the legend appearance instead of adding a separate `guides()` layer.  The `guide` argument is part of all scale functions.

For example, let's say I was going to use `scale_color_viridis_c()` to change the default color palette of the whole plot (i.e., plot and legend).  Then I use the same `guide_legend()` code from above for the `guide` argument.  HEre I again change the size of the points in the legend.  

```{r firstplot3}
ggplot(data = diamonds, aes(x = carat, y = price, color = cut) ) +
     geom_point(alpha = .25, size = 1) +
     scale_color_viridis_d(option = "A",
                           guide = guide_legend(override.aes = list(size = 3) ) )

```

## Changing multiple aesthetic parameters

You can control multiple aesthetic parameters at once by adding them to the list passed to `override.aes`.  If I want to increase the point size as well as remove the point transparency, I can change both `size` and `alpha`.

```{r firstplot4}
ggplot(data = diamonds, aes(x = carat, y = price, color = cut) ) +
     geom_point(alpha = .25, size = 1) +
     scale_color_viridis_d(option = "A",
                           guide = guide_legend(override.aes = list(size = 3,
                                                                    alpha = 1) ) )

```

# Suppress aesthetics from part of the legend

Removing aesthetic for only some parts of the legend is another common use for `override.aes`.  For example, this can be useful when different layers are based on different amounts of data.  

This following example is based on [this Stack Overflow question](https://stackoverflow.com/questions/59548358/r-ggplot2-in-the-legend-how-do-i-hide-unused-colors-from-one-geom-while-show).  You can see the points layer has data based on all three groups used for the `color` aesthetic but the rectangle is drawn for only a single group.  Colored outlines, representing the rectangles, are added to the legend for every group. 

```{r secondplot}
points = structure(list(x = c(5L, 10L, 7L, 9L, 86L, 46L, 22L, 94L, 21L, 
6L, 24L, 3L), y = c(51L, 54L, 50L, 60L, 97L, 74L, 59L, 68L, 45L, 
56L, 25L, 70L), id = c("a", "a", "a", "a", "b", "b", "b", "b", 
"c", "c", "c", "c")), row.names = c(NA, -12L), class = "data.frame")
box = data.frame(left = 1, right = 10, bottom = 50, top = 60, id = "a")

ggplot(data = points, aes(color = id) ) +
     geom_point(aes(x = x, y = y) ) +
     geom_rect(data = box, aes(xmin = left,
                               xmax = right,
                               ymin = 50,
                               ymax = top),
               fill = NA)

```

Those outlines are based on lines, which is the `linetype` aesthetic.  Removing the outlines for the second and third groups would make the legend represent what is shown in the plot better.  Suppressing these lines can be done with `override.aes`, setting the lines to `0` in order to remove them.  

Note that I have to list the line for every group, not just the groups I want to remove.  I keep the line for the first group solid by using a `1`.

```{r secondplot2}
ggplot(data = points, aes(color = id) ) +
     geom_point(aes(x = x, y = y) ) +
     geom_rect(data = box, aes(xmin = left,
                               xmax = right,
                               ymin = 50,
                               ymax = top),
               fill = NA) +
     guides(color = guide_legend(override.aes = list(linetype = c(1, 0, 0) ) ) )

```

# Combined legend

Legend based on constants to different aesthetics

https://stackoverflow.com/questions/38838349/ggplot2-adjust-legend-symbols-in-overlayed-plot/38838895#38838895
https://stackoverflow.com/questions/53118540/ggplot2-solid-line-for-one-group-points-for-the-other/53119159#53119159
https://stackoverflow.com/questions/52740583/add-legend-for-a-custom-stat-summary-fun-data/52745072#52745072

Hline version
https://stackoverflow.com/questions/39119917/how-to-add-a-legend-to-hline/39127187#39127187

Combined variable for two different aesthetics in one legend

https://stackoverflow.com/questions/34318292/controlling-legend-labels-for-a-plot-with-color-and-linetype-aesthetic-attribute/34319103#34319103

# Change shape for fill legend?

https://stackoverflow.com/questions/44765946/r-ggplot2-plot-geom-point-with-black-and-white-without-using-shape/44766180#44766180

# Just the code, please

```{r getlabels, echo = FALSE, purl = FALSE}
labs = knitr::all_labels()
labs = labs[!labs %in% c("setup", "toc", "getlabels", "allcode", "makescript")]
```

```{r makescript, include = FALSE, purl = FALSE}
options(knitr.duplicate.label = "allow") # Needed to purl like this
input  = knitr::current_input()  # filename of input document
scriptpath = paste(tools::file_path_sans_ext(input), "R", sep = ".")
output = here::here("static", "script", scriptpath)
knitr::purl(input, output, documentation = 0, quiet = TRUE)
```

Here's the code without all the discussion.  Copy and paste the code below or you can download an R script of uncommented code [from here](`r paste0("/script/", scriptpath)`).

```{r allcode, ref.label = labs, eval = FALSE, purl = FALSE}
```
